---
title: "labwork_tidycensus"
author: "Bayowa"
date: "2024-11-13"
output: github_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidycensus)

# Set your Census API key
census_api_key("e06a3ef2522b01fbf59f0fb986f724a11ed67dd2",overwrite= TRUE)

# View available variables for 5-year ACS data
acs_variables <- load_variables(year = 2022, dataset = "acs5")
print(acs_variables)
View(acs_variables)
```




# Block to get city details using city codes,variables, data year
```{r}
# Load necessary libraries
library(tidycensus)
library(dplyr)

# Set your Census API key
census_api_key("e06a3ef2522b01fbf59f0fb986f724a11ed67dd2", overwrite = TRUE)

# Define a helper function to handle state abbreviations, full names, and FIPS codes
get_state_fips <- function(input_states) {
  # Data frame of state names, abbreviations, and FIPS codes
  state_info <- data.frame(
    state_name = c(state.name, "District of Columbia"),
    state_abb = c(state.abb, "DC"),
    fips_code = sprintf("%02d", c(1:51)) # FIPS codes 01 to 51
  )
  
  # Identify if input is FIPS, abbreviation, or full name, and return FIPS code
  fips_codes <- sapply(input_states, function(state) {
    if (state %in% state_info$fips_code) {
      return(state)  # Already a FIPS code
    } else if (state %in% state_info$state_abb) {
      return(state_info$fips_code[state_info$state_abb == state])
    } else if (state %in% state_info$state_name) {
      return(state_info$fips_code[state_info$state_name == state])
    } else {
      stop("Invalid state input: ", state)
    }
  })
  
  return(fips_codes)
}

# Define the states you're interested in, using FIPS codes, abbreviations, or full names
input_states <- c("36", "SC", "Virginia", "48") # Example FIPS, abbreviation, and full name

# Convert input to FIPS codes
fips_states <- get_state_fips(input_states)

# Define the range of years, e.g., 2010 to 2022
years <- 2010:2022

# Initialize an empty dataframe to store results
black_data <- data.frame()

# Loop through each year and get the ACS data for Black median income and Black median age
for (year in years) {
  
  # Skip the year 2020
  if (year == 2020) {
    next  # Skip this iteration of the loop
  }
  
  # Get median household income for Black population
  income_data <- get_acs(geography = "state", 
                         variables = "B19013B_001",  # Median household income for Black
                         state = fips_states, 
                         year = year, 
                         survey = "acs1")  # 1-year estimates for each year
  
  # Add the 'year' column to income data
  income_data <- income_data %>%
    mutate(year = year)  # Add the current year to the dataset
  
  # Get median age for Black population
  age_data <- get_acs(geography = "state", 
                      variables = "B01002B_001",  # Median age for Black
                      state = fips_states, 
                      year = year, 
                      survey = "acs1")
  
  # Add the 'year' column to age data
  age_data <- age_data %>%
    mutate(year = year)  # Add the current year to the dataset
  
  # Combine both datasets for that year
  combined_data <- income_data %>%
    rename(median_income = estimate) %>%
    left_join(age_data %>% rename(median_age = estimate), by = c("GEOID", "NAME", "year"))
  
  # Add the combined data to the overall dataframe
  black_data <- bind_rows(black_data, combined_data)
}

# View the resulting dataset with all the required years' data
black_data
```


# Block of code to get city details using specific city fips code,abbreviation, variables and dataset year
```{r}
# Load necessary libraries
library(tidycensus)
library(dplyr)

# Set your Census API key
census_api_key("e06a3ef2522b01fbf59f0fb986f724a11ed67dd2", overwrite = TRUE)

# Load the correct state information from tidycensus package
state_info <- tigris::fips_codes %>%
  distinct(state, state_code, state_name) %>%
  rename(state_abb = state, fips_code = state_code)

# Function to get state abbreviation from FIPS code or state abbreviation
get_state_abbreviation <- function(input_code) {
  # Check if input is a FIPS code
  if (input_code %in% state_info$fips_code) {
    state_row <- state_info[state_info$fips_code == input_code, ]
  } else if (input_code %in% state_info$state_abb) {  # Check if input is a state abbreviation
    state_row <- state_info[state_info$state_abb == input_code, ]
  } else {
    stop("Invalid FIPS code or state abbreviation provided.")
  }
  
  return(state_row$state_abb)  # Return the state abbreviation
}

# Function to get data for a specific year, FIPS code, or state abbreviation, and variable(s)
get_data_for_year_state_vars <- function(year_input, input_code, variables) {
  
  # Get the state abbreviation from either FIPS code or state abbreviation
  state_input <- get_state_abbreviation(input_code)
  
  # Get data for specified variables
  data <- get_acs(geography = "state",
                  variables = variables,
                  state = state_input,
                  year = year_input,
                  survey = "acs1")
  
  # Add the year column to the data
  data <- data %>%
    mutate(year = year_input)
  
  return(data)
}

# Input FIPS code or state abbreviation and year here
year_input <- 2014
input_code <- "SC"  # FIPS code or state abbreviation here

# Get median household income and median age for Black population
variables <- c("B19013B_001", "B01002B_001")
data <- get_data_for_year_state_vars(year_input, input_code, variables)

print(data)

```


# Block to get specific variable details using var and year
```{r}
# Load necessary libraries
library(tidycensus)
library(dplyr)

# Set your Census API key
census_api_key("e06a3ef2522b01fbf59f0fb986f724a11ed67dd2", overwrite = TRUE)

# Function to retrieve variable details by inputting variable code
get_variable_details <- function(variable_code, year = 2020, dataset = "acs1") {
  
  # Load all available variables for the specified year and dataset
  variables <- load_variables(year = year, dataset = dataset)
  
  # Filter for the specific variable code
  variable_details <- variables %>%
    filter(name == variable_code)
  
  # Check if the variable exists
  if (nrow(variable_details) == 0) {
    message("Variable not found. Please check the variable code and try again.")
  } else {
    return(variable_details)
  }
}

#input var here
variable_code <- "B01002B_001"  # Replace with any variable code you want details for
year <- 2019  # Specify the year of interest
dataset <- "acs1"  # Specify the dataset (e.g., "acs1" for 1-year ACS estimates)

# Get details for the specified variable
variable_info <- get_variable_details(variable_code, year, dataset)
print(variable_info)

```


BLock of code to get table needed with variable details
```{r}
# Load necessary libraries
library(tidycensus)
library(dplyr)

# Set your Census API key
census_api_key("e06a3ef2522b01fbf59f0fb986f724a11ed67dd2", overwrite = TRUE)

# Function to retrieve variable details and data for a specific year, FIPS code or state abbreviation, and variable(s)
get_data_with_variable_details <- function(year_input, input_code, variables, dataset = "acs1") {
  
  # Load variable details for the specified year and dataset
  variable_details <- load_variables(year = year_input, dataset = dataset) %>%
    filter(name %in% variables)
  
  # Check if any variables were not found
  if (nrow(variable_details) == 0) {
    stop("One or more variables not found. Please check the variable codes.")
  }
  
  # Get the data for specified variables
  data <- get_acs(geography = "state",
                  variables = variables,
                  state = input_code,
                  year = year_input,
                  survey = dataset) %>%
    mutate(year = year_input)  # Add the year column
  
  # Merge the variable details with the ACS data
  enriched_data <- data %>%
    left_join(variable_details, by = c("variable" = "name")) %>%
    select(GEOID, NAME, year, variable, label, concept, estimate, moe)
  
  return(enriched_data)
}

# Input your year, FIPS code or state abbreviation, and variable codes here
year_input <- 2022
input_code <- "SC"  # You can use a FIPS code or state abbreviation
variables <- c("B19013B_001", "B01002B_001")  # Median household income for Black and median age for Black

# Retrieve data with variable labels and concepts
data_with_details <- get_data_with_variable_details(year_input, input_code, variables)
print(data_with_details)
```

#Block of code to get table with variable details using year, state fips code/abbreviation and var
```{r}
# Load necessary libraries
library(tidycensus)
library(dplyr)

# Set your Census API key
census_api_key("e06a3ef2522b01fbf59f0fb986f724a11ed67dd2", overwrite = TRUE)

# Function to retrieve variable details and data for a specific year, FIPS code or state abbreviation, and variable(s)
get_data_with_variable_details <- function(year_input, input_code, variables, dataset = "acs1") {
  
  # Load variable details for the specified year and dataset
  variable_details <- load_variables(year = year_input, dataset = dataset) %>%
    filter(name %in% variables)
  
  # Check if any variables were not found
  if (nrow(variable_details) == 0) {
    stop("One or more variables not found. Please check the variable codes.")
  }
  
  # Get the data for specified variables
  data <- get_acs(geography = "state",
                  variables = variables,
                  state = input_code,
                  year = year_input,
                  survey = dataset) %>%
    mutate(year = year_input)  # Add the year column
  
  # Merge the variable details with the ACS data
  enriched_data <- data %>%
    left_join(variable_details, by = c("variable" = "name")) %>%
    select(NAME, year, label, concept, estimate, moe)  # Exclude variable and GEOID columns here
  
  return(enriched_data)
}

# Input your year, FIPS code or state abbreviation, and variable codes here
year_input <- 2022
input_code <- c("SC", "NC", "MD")  # You can use a FIPS code or state abbreviation
variables <- c("B19013B_001", "B01002B_001")  # Median household income for Black and median age for Black

# Retrieve data with variable labels and concepts
data_with_details <- get_data_with_variable_details(year_input, input_code, variables)
print(data_with_details)
```


#Block to save as csv file for use in any IDE
```{r}
write.csv(data_with_details, file = "data_with_details_output.csv", row.names = FALSE)
```

